Fast and adaptive cointegration based model for forecasting financial time series

### 2017-12-10 slides/defense-slides.tex
nocite references, first contribution ready

\section{Contributions}
\begin{frame}
\frametitle{First contribution: Adaptive VECM}
\begin{figure}
\includegraphics[width=\textwidth]{img/slidingwindow}
\end{figure}
\end{frame}

\begin{frame}
\frametitle{First approach idea}
An adaptive Vector Error Correction (AVECM) algorithm that:
\begin{itemize}
\item considers the last most recent data (windows size L)
\item finds optimal parameters of VECM in that window.
\item the optimisation is made in a distributed environment.
\end{itemize}
\end{frame}

@@

\begin{frame}
\frametitle{Percentage of cointegration}
Since $r=0$ means no cointegration and $r=l=4$ reveals that no process is I(1) but stationary.
The interesting cases of cointegration are those where $r$ lies strictly
between $0$ and $l$, i.e. $0<r<l$.
\begin{block}{Percentage of cointegration (PC)}
{\color{blue}
\begin{equation*} \label{eq:pcoint}
PC = 
\frac{\#\{ it \mid \text{$it$ has $r$ c.v. with $0<r<l$}\}}
     {\#it}\times 100
\end{equation*}}
\end{block}
\end{frame}

@@

\begin{frame}
\frametitle{PC vs MSE}
\begin{figure}[ht!]
  %\vspace{-0.8cm}
  \centering
  \includegraphics[width=0.7\textwidth]{img/51_Fig2}
  \caption{MSE versus the percentage of cointegration considering 1000
  iterations. }
  \label{fig:cointvsmse}
\end{figure}
\end{frame}
%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\begin{frame}
\frametitle{Adaptive VECM algorithm (AVECM)}
\begin{itemize}
\item We proposed to choose $L$ and $p$ in order to maximise the percentage of
cointegration $PC$ in the near past. This process was done every time that new data was processed. 
\item To use a distributed environment to make this parameters search, since this is the most expensive routine.
\end{itemize}
\end{frame}

@@

%\subsection{Model comparison} 
%\begin{frame}
%\frametitle{Random walk}
%The random walk model is defined as:
%\begin{equation}
%\mathbf{y}_t = \mathbf{y}_{t-1} + \epsilon_{t}
%\label{rwmodel}
%\end{equation}
%The naive forecast of the time series difference $\hat{\mathbf{y}}_{t+1}$ for the random walk model is defined as:
%\begin{equation}
%\hat{\mathbf{y}}_{t+1} = \mathbf{y}_t + \hat{\epsilon}_{t+1} 
%\end{equation}
%\noindent where  $\hat{\epsilon}_{t+1} = \epsilon_{t}$.
%\end{frame}
%
%
%\begin{frame}
%\frametitle{ARIMA}
%A process can be modelled as an ARIMA$(p,d,q)$ if $\mathbf{x}_t = \Delta^d \mathbf{y}_t $, is an ARMA$(p,q)$. An ARMA$(p,q)$ model is the following:
%\begin{equation}
%\mathbf{x}_t = \sum_{i=1}^p \phi_i \mathbf{x}_{t-i}  +  \sum_{j=1}^q \theta_j \epsilon_{t-j}  
%\end{equation}
%\noindent with coefficients $\phi_p \neq 0$, $\theta_q \neq 0$ and $\sigma_{\epsilon}^2 > 0$.
%\end{frame}

%%\begin{frame}
%%\frametitle{Evaluation methods}
%%\begin{block}{MSE}
%%Measures the distance between forecasts
%%and the true values and large deviations from the true value have a
%%large impact due to squaring forecast error.
%%{\color{blue}
%%\begin{equation}\label{eq:MSE}
%%\text{MSE} = 
%%\frac{\displaystyle \sum_{t=1}^{N} (\mathbf{y}_t-\hat{\mathbf{y}}_t)^2}{N}
%%\end{equation}}
%%\end{block}
%%\begin{block}{\bf $U$-statistic}
%%Unit free measure obtained as the ratio between the root MSE (RMSE) of a model and the RMSE of the naive random walk model. 
%%\end{block}
%%\end{frame}
%
%\section{Experiments}
%\subsection{Uno}
%
\begin{frame}
\frametitle{Unit root tests}
VECM can only be used if the time series are I(1). Therefore, 
before running the tests, we firstly checked whether the time series were
I(1) using the Augmented Dickey Fuller (ADF) test with lags $p=1,2,3,4,5$.
\begin{table}[ht]
\centering
\tiny
\begin{tabular}{llllll}
\toprule
{Variable} & {ADF(1)} & {ADF(2)} & {ADF(3)} & {ADF(4)} & {ADF(5)}\\ 
\midrule
EURUSD &  -0.052   & -0.054  & -0.054  & -0.054  & -0.054  \\
GBPUSD &  -0.744  & -0.784  & -0.805  & -0.837  & -0.846  \\
USDCHF &  -0.476   & -0.493  & -0.493  & -0.495  & -0.502  \\
USDJPY &  0.357   & 0.360  & 0.360  & 0.367  & 0.367  \\
$\Delta$ EURUSD & -128.4*  & -128.4*  & -96.85* & -89.12*   & -89.12*\\
$\Delta$ GBPUSD & -131.4*  & -112.7*  & -102.5* & -92.86*   & -88.29*\\
$\Delta$ USDCHF & -127.8*  & -127.8*  & -96.94* & -88.82*   & -80.79*\\
$\Delta$ USDJPY & -135.1*  & -135.1*  & -101.2* & -101.2*   & -101.2*\\
\bottomrule
\addlinespace[1ex]
\multicolumn{6}{l}{ \textsuperscript{*} Indicates significance at 1\% level}
\end{tabular}
\caption{Unit roots tests for EURUSD, GBPUSD, USDCHF and USDJPY at 10-second
frequency.}
\label{tab:adf}
\end{table}
\end{frame}

@@

\hline
 EURUSD & 1.0702 e-09 & 1.1481 e-09 &  9.250 e-12 & & 0.6863 & 0.7108\\
 GBPUSD & 1.6630 e-09 & 1.7408 e-09 &  6.951 e-02 & & 0.6866 & 0.7025\\
 USDCHF & 5.8503 e-10 & 6.3545 e-10 &  2.899 e-14 & & 0.6803 & 0.7091\\
 USDJPY  & 6.3483 e-06 & 6.5194 e-06 &  6.853 e-05 & & 0.6964 & 0.7057
 \end{tabular}
 \end{frame}
%%For AVECM we considered different number of iterations (parameter $m$ in algorithm \ref{alg:AVECM}): 10, 50 and 100. We tried 12, 24 and 47 different pair of combinations for $L$ and $p$. Possible values for $L$ were in the interval $[100,4000]$ and $p$ can have values in between $[1,5]$. Best AVECM performance was compared against ARIMA and the random walk model.
%%Table~\ref{tab:stats} shows the out-of-sample performance measures: MSE and $U$-statistic for AVECM and ARIMA. In terms of both measures we found that AVECM is superior to ARIMA and the naive random walk model. We also included the p-value that proves that the difference in the MSE is significant at the 99\% significance level in three of the four currency rates and at 90\% in the case of GBPUSD. The $U$-statistic shows that AVECM and ARIMA are superior to the naive random walk model and that our proposal is also superior to ARIMA.
%
%
%
\begin{frame}
\frametitle{Parallel implementation}
The function that get best parameters $L$ and $p$ was implemented using MPI in Python. MPI allows large-scale parallel applications
with wide portability to be built, being able to run in large clusters
or on local computers.
Tests ran in a cluster with 2 servers Xeon E5-2667 (2.90GHz)
of 24 cores each (48 cores in total) and 24GB RAM.
Parameter settings:
\begin{itemize}
%\item Parameter $it=100$
\item The $L$ parameter was always chosen between 100 and 4000 and $p$ always took values
between 1 and 5. 
\item Parameter $nparams$ represents the number of pairs ($L$,$p$) used to maximise the percentage of cointegration. 
\end{itemize}
\end{frame}
%
\begin{frame}
\frametitle{Execution times}
\begin{figure}[ht]
  \centering
  \includegraphics[width=0.7\textwidth]{img/51_Fig3}
  \caption[Computing time and Speed-up]{Computing time of sequential and parallel algorithm is shown in the
  upper figure. Speed-up is shown below.}
  \label{fig:extimes}
\end{figure}
\end{frame}

\begin{frame}
\frametitle{Conclusions}
\begin{itemize}
\item Cointegration relations change with time. We empirically showed that the Johansen method is sensitive to the number of lags but also to the amount of data considered.
\item We found that out-of-sample forecast performance MSE is related to the {\em percentage of cointegration\/}.  
\item VECM improves performance measures by finding
parameters of $L$ and $p$ maximising the percentage of cointegration.
\item Increasing the number of parameters will always lead to
better performance measures.
\item The parallel implementation allowed the execution times to be reduced
more than 9 times and therefore a response time was obtain before 10
seconds. 
\end{itemize}
\end{frame}

@@

@@

@@

@@

\bibliographystyle{amscls/amsalpha}
\bibliography{reference}
\end{document}

>>>>

\section{Contributions}
\begin{frame}
\frametitle{First contribution: Adaptive VECM}
\hspace*{-9mm}
\includegraphics[width=1.15\textwidth]{img/slidingwindow}
\end{frame}

@@

\begin{frame}
\frametitle{Percentage of cointegration}
Since $r=0$ means no cointegration and $r=l=4$ reveals that no process is I(1) but stationary.
The interesting cases of cointegration are those where $r$ lies strictly
between $0$ and $l$, i.e. $0<r<l$.
\vspace{5mm}
\begin{block}{Percentage of cointegration (PC)}
{\color{blue}
\begin{equation*} \label{eq:pcoint}
PC = 
\frac{\#\{ it \mid \text{$it$ has $r$ c.v. with $0<r<l$}\}}
     {\#it}\times 100
\end{equation*}}
\end{block}
\end{frame}

@@

\begin{frame}
\frametitle{PC vs MSE}
\begin{figure}[ht!]
  %\vspace{-0.8cm}
  \centering
  \includegraphics[width=0.7\textwidth]{img/51_Fig2}
  \caption{MSE versus the percentage of cointegration considering 1000
  iterations. }
  \label{fig:cointvsmse}
\end{figure}
\end{frame}

\begin{frame}
\frametitle{Adaptive VECM algorithm (AVECM)}
AVECM has the following features: 
\begin{enumerate}
\item Uses VECM only considering a sliding windows of size $L$ of data.
\item We proposed to choose $L$ and the number of lags $p$ in order to maximise the percentage of cointegration $PC$ in the near past. This process was done every time that new data was processed. 
\item To use a distributed environment to make this parameters search, since this is the most expensive routine.
\end{enumerate}
\end{frame}

@@

%%\begin{frame}
%%\frametitle{Evaluation methods}
%%\begin{block}{MSE}
%%Measures the distance between forecasts
%%and the true values and large deviations from the true value have a
%%large impact due to squaring forecast error.
%%{\color{blue}
%%\begin{equation}\label{eq:MSE}
%%\text{MSE} = 
%%\frac{\displaystyle \sum_{t=1}^{N} (\mathbf{y}_t-\hat{\mathbf{y}}_t)^2}{N}
%%\end{equation}}
%%\end{block}
%%\begin{block}{\bf $U$-statistic}
%%Unit free measure obtained as the ratio between the root MSE (RMSE) of a model and the RMSE of the naive random walk model. 
%%\end{block}
%%\end{frame}
%
%\section{Experiments}
%\subsection{Uno}
%
\begin{frame}
\frametitle{Unit root tests}
Augmented Dickey Fuller (ADF) test with lags $p=1,2,3,4,5$.
\begin{table}[ht]
\centering
\small
\begin{tabular}{llllll}
\toprule
{Variable} & {ADF(1)} & {ADF(2)} & {ADF(3)} & {ADF(4)} & {ADF(5)}\\ 
\midrule
EURUSD &  -0.052   & -0.054  & -0.054  & -0.054  & -0.054  \\
GBPUSD &  -0.744  & -0.784  & -0.805  & -0.837  & -0.846  \\
USDCHF &  -0.476   & -0.493  & -0.493  & -0.495  & -0.502  \\
USDJPY &  0.357   & 0.360  & 0.360  & 0.367  & 0.367  \\
$\Delta$ EURUSD & -128.4*  & -128.4*  & -96.85* & -89.12*   & -89.12*\\
$\Delta$ GBPUSD & -131.4*  & -112.7*  & -102.5* & -92.86*   & -88.29*\\
$\Delta$ USDCHF & -127.8*  & -127.8*  & -96.94* & -88.82*   & -80.79*\\
$\Delta$ USDJPY & -135.1*  & -135.1*  & -101.2* & -101.2*   & -101.2*\\
\bottomrule
\addlinespace[1ex]
\end{tabular}
\caption{Unit roots tests for EURUSD, GBPUSD, USDCHF and USDJPY at 10-second
frequency.}
\label{tab:adf}
\end{table}
\end{frame}

@@

\hline
 EURUSD & 1.0702 e-09 & 1.1481 e-09 &  9.250 e-12 & & 0.6863 & 0.7108\\
 GBPUSD & 1.6630 e-09 & 1.7408 e-09 &  6.951 e-02 & & 0.6866 & 0.7025\\
 USDCHF & 5.8503 e-10 & 6.3545 e-10 &  2.899 e-14 & & 0.6803 & 0.7091\\
 USDJPY  & 6.3483 e-06 & 6.5194 e-06 &  6.853 e-05 & & 0.6964 & 0.7057
 \end{tabular}
 \end{frame}
%%For AVECM we considered different number of iterations (parameter $m$ in algorithm \ref{alg:AVECM}): 10, 50 and 100. We tried 12, 24 and 47 different pair of combinations for $L$ and $p$. Possible values for $L$ were in the interval $[100,4000]$ and $p$ can have values in between $[1,5]$. Best AVECM performance was compared against ARIMA and the random walk model.
%%Table~\ref{tab:stats} shows the out-of-sample performance measures: MSE and $U$-statistic for AVECM and ARIMA. In terms of both measures we found that AVECM is superior to ARIMA and the naive random walk model. We also included the p-value that proves that the difference in the MSE is significant at the 99\% significance level in three of the four currency rates and at 90\% in the case of GBPUSD. The $U$-statistic shows that AVECM and ARIMA are superior to the naive random walk model and that our proposal is also superior to ARIMA.
%
%
%
\begin{frame}
\frametitle{Parallel implementation}
The function that get optimal parameters $L$ and $p$ was implemented using MPI in Python. Tests ran in a cluster with 2 servers Xeon E5-2667 (2.90GHz) of 24 cores each (48 cores in total) and 24GB RAM.

Parameter settings:
\begin{itemize}
%\item Parameter $it=100$
\item The $L$ parameter was always chosen between 100 and 4000 and $p$ always took values between 1 and 5. 
\item Parameter $nparams$ represents the number of pairs ($L$,$p$) used to maximise the percentage of cointegration. 
\end{itemize}
\end{frame}
%
\begin{frame}
\frametitle{Execution times}
\begin{figure}[ht]
  \centering
  \includegraphics[width=0.9\textwidth]{img/51_Fig3}
  \caption[Computing time and Speed-up]{Computing time of sequential and parallel algorithm is shown in the
  upper figure. Speed-up is shown below.}
  \label{fig:extimes}
\end{figure}
\end{frame}

\begin{frame}
\frametitle{Conclusions}
\begin{itemize}
\item Cointegration relations change with time. We empirically showed that the Johansen method is sensitive to the number of lags but also to the amount of data considered.
%\item We found that out-of-sample forecast performance MSE is related to the {\em percentage of cointegration\/}.  
\item VECM improves performance measures by finding
parameters of $L$ and $p$ maximising the percentage of cointegration.
%\item Increasing the number of parameters will always lead to
%better performance measures.
\item The parallel implementation allowed the execution times to be reduced
more than 9 times and therefore a response time was obtain before 10
seconds. 
\end{itemize}
\end{frame}

@@

\nocite{arce+salinas2012,icpram15,Arce2017}
\bibliographystyle{apalike}
\bibliography{reference}

@@

\appendix

@@

%\subsection{Model comparison} 
\begin{frame}
\frametitle{Random walk}
The random walk model is defined as:
\begin{equation}
\mathbf{y}_t = \mathbf{y}_{t-1} + \epsilon_{t}
\label{rwmodel}
\end{equation}
The naive forecast of the time series difference $\hat{\mathbf{y}}_{t+1}$ for the random walk model is defined as:
\begin{equation}
\hat{\mathbf{y}}_{t+1} = \mathbf{y}_t + \hat{\epsilon}_{t+1} 
\end{equation}
\noindent where  $\hat{\epsilon}_{t+1} = \epsilon_{t}$.
\end{frame}
%
%
\begin{frame}
\frametitle{ARIMA}
A process can be modelled as an ARIMA$(p,d,q)$ if $\mathbf{x}_t = \Delta^d \mathbf{y}_t $, is an ARMA$(p,q)$. An ARMA$(p,q)$ model is the following:
\begin{equation}
\mathbf{x}_t = \sum_{i=1}^p \phi_i \mathbf{x}_{t-i}  +  \sum_{j=1}^q \theta_j \epsilon_{t-j}  
\end{equation}
\noindent with coefficients $\phi_p \neq 0$, $\theta_q \neq 0$ and $\sigma_{\epsilon}^2 > 0$.
\end{frame}

@@

\end{document}
